Meta-analysis on diagnosis in MSSM-Penn-Pitt and HBCC ACC/DLPFC

### 2020-07-31 individual-cohorts/meta-analysis.Rmd
metaanalysis by diagnosis

---
title: "Sex-specific Meta-analysis in MSSM-Penn-Pitt and HBCC ACC/DLPFC"
author: "KM and GH"
date: "`r date()`"
output: html_document
---

```{r knit2synapse, eval=FALSE}
# rmarkdown::render("graphics_t_statistics.Rmd")
library(synapser)
library(knit2synapse)
synLogin()
knit2synapse::createAndKnitToFolderEntity(file = "individual-cohorts/graphics_t_statistics.Rmd",
                                          parentId ="syn21435176",
                                          folderName = 'Sex-specific Meta-analysis')
```

```{r setup, include=FALSE}
library(synapser)
synLogin()
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(ComplexHeatmap)
library(readxl)
library(gtools)
library(corrplot)
library(pvclust)
library(metafor)

@@

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=FALSE, eval=FALSE}
thisFileName <- 'graphics_t_statistics.Rmd'

@@

@@

## t-statistics from differential expression results

- separate by comparison

```{r data}
mpp_acc <- readr::read_tsv(synGet("syn22079117")$path) %>% 
  dplyr::select(Comparison, gene_id, AveExpr, t, logFC, P.Value, adj.P.Val) %>% 
  dplyr::rename(`MSSM-Penn-Pitt ACC t-statistic` = t)
mpp_dlpfc <- readr::read_tsv(synGet("syn22079472")$path) %>% 
  dplyr::select(Comparison, gene_id, AveExpr, t, logFC, P.Value, adj.P.Val) %>% 
  dplyr::rename(`MSSM-Penn-Pitt DLPFC t-statistic` = t)
hbcc_dsACC <- readr::read_tsv(synGet("syn22086829")$path) %>% 
  dplyr::select(Comparison, gene_id, AveExpr, t, logFC, P.Value, adj.P.Val) %>% 
  dplyr::rename(`HBCC dsACC t-statistic` = t)
hbcc_stg <- readr::read_tsv(synGet("syn22087386")$path) %>% 
  dplyr::select(Comparison, gene_id, AveExpr, t) %>% 
  dplyr::rename(`HBCC STG t-statistic` = t)
hbcc_sgACC <- readr::read_tsv(synGet("syn22085862")$path) %>% 
  dplyr::select(Comparison, gene_id, AveExpr, t, logFC, P.Value, adj.P.Val) %>% 
  dplyr::rename(`HBCC sgACC t-statistic` = t)
hbcc_dlpfc <- readr::read_tsv(synGet("syn22084205")$path) %>% 
  dplyr::select(Comparison, gene_id, AveExpr, t, logFC, P.Value, adj.P.Val) %>% 
  dplyr::rename(`HBCC DLPFC t-statistic` = t)
# CMC Analysis 1
CMC1 <- readr::read_tsv(synGet("syn5607603")$path) %>% 
  dplyr::rename(gene_id = genes,
                `CMC 1 t-statistic` = t) %>% 
  dplyr::select(gene_id, `CMC 1 t-statistic`)
# HBCC-DLPFC microarray
CMC_HBCC <- readr::read_tsv(synGet("syn4941553")$path) %>% 
  dplyr::rename(gene_id = genes,
                `HBCC microarray t-statistic` = t) %>% 
  dplyr::select(gene_id, `HBCC microarray t-statistic`)
# Gandal, Science 2018, Table S1
# Use Tab:DGE
# wget https://science.sciencemag.org/highwire/filestream/720110/field_highwire_adjunct_files/1/aat8127_Table_S1.xlsx
gandal <- readxl::read_xlsx( synGet("syn22255217")$path, sheet = "DGE") %>% 
  dplyr::rename(gene_id = ensembl_gene_id,
                `Gandal, SCZ t-statistic` = `SCZ.t.value`,
                `Gandal, BD t-statistic` = `BD.t.value`,
                `Gandal, ASD t-statistic` = `ASD.t.value`) %>%
  dplyr::select(gene_id, `Gandal, SCZ t-statistic`,
                `Gandal, BD t-statistic`, `Gandal, ASD t-statistic`)

# AMP-AD RNASeq reprocessing GRCh37 - https://www.synapse.org/#!Synapse:syn9702085
MSSM <- readr::read_tsv(synGet("syn10526259")$path) %>%
  dplyr::filter(Comparison %in% "AD-CONTROL",
                Model %in% "Diagnosis") %>%
  dplyr::select(ensembl_gene_id, Tissue, t) %>%
  dplyr::group_by(ensembl_gene_id, Tissue) %>% 
  dplyr::filter(row_number() == 1) %>% 
  tidyr::pivot_wider(names_from = Tissue,
                     values_from = t) %>%
  dplyr::rename(`AMP-AD MSSM FP AD t-statistic` = FP,
                `AMP-AD MSSM IGF AD t-statistic` = IFG,
                `AMP-AD MSSM PHG AD t-statistic` = PHG,
                `AMP-AD MSSM STG AD t-statistic` = STG,
                gene_id = ensembl_gene_id) %>% 
  dplyr::select(gene_id, `AMP-AD MSSM STG AD t-statistic`) 
MSSM <- MSSM[!is.na(MSSM$`AMP-AD MSSM STG AD t-statistic`),]

MAYO <- readr::read_tsv(synGet("syn8468023")$path) %>% 
  dplyr::filter(Comparison %in% "AD-CONTROL",
                Model %in% "Diagnosis") %>%
  dplyr::select(ensembl_gene_id, `Tissue.ref`, t) %>%
  dplyr::group_by(ensembl_gene_id, `Tissue.ref`) %>% 
  dplyr::filter(row_number() == 1) %>% 
  tidyr::pivot_wider(names_from = `Tissue.ref`,
                     values_from = t) %>% 
  dplyr::rename(`AMP-AD MAYO CBE AD t-statistic` = CBE,
                `AMP-AD MAYO TCX AD t-statistic` = TCX,
                gene_id = ensembl_gene_id) %>% 
  dplyr::select(-`AMP-AD MAYO TCX AD t-statistic`)

ROSMAP <- readr::read_tsv(synGet("syn8456721")$path) %>%
    dplyr::filter(Comparison %in% "AD-CONTROL",
                Model %in% "Diagnosis") %>%
  dplyr::rename(gene_id = ensembl_gene_id, 
                `AMP-AD ROSMAP DLPFC AD t-statistic` = t) %>% 
  dplyr::select(gene_id, `AMP-AD ROSMAP DLPFC AD t-statistic`)

out_group <- inner_join(MSSM, MAYO) %>% 
  inner_join(., ROSMAP) %>% 
  inner_join(., gandal)

df <- inner_join(mpp_acc, mpp_dlpfc,
                 by = c("Comparison", "gene_id"))
df <- inner_join(df, hbcc_dsACC, by = c("Comparison", "gene_id"))
df <- inner_join(df, hbcc_stg, by = c("Comparison", "gene_id"))
df <- inner_join(df, hbcc_sgACC, by = c("Comparison", "gene_id"))
df <- inner_join(df, hbcc_dlpfc, by = c("Comparison", "gene_id"))
```

```{r functions}
plot_t <- function(df, conditions){
  p = list()
  for(i in seq_along(conditions[,1])){
  positions = conditions[i,]
  x = df[positions[1]]
  y = df[positions[2]]
  computed_cor = cor(x, y, method = "spearman")
  x_var <- names(x)
  y_var <- names(y)
  p[[i]] = ggplot(df, aes(x = .data[[x_var]], y = .data[[y_var]])) +
  geom_hex() +
  geom_abline() +
  geom_smooth(method = "lm") +
  theme_bw() +
  xlim(8,-8) + 
  ylim(8,-8) +
    labs(
    caption = paste0("spearman correlation between the statistics is ", round(computed_cor, digits = 3))
  ) +
    theme(aspect.ratio=1, 
          axis.text=element_text(size=12),
          plot.caption = element_text(face = "bold", size = "12"))
  }
multiplot(plotlist = p, cols = 2)
}
```

# Control vs Other
```{r plota, fig.width = 10, fig.height = 100}
df_one <- df[df$Comparison %in% "Control_vs_Other",] %>% 
  dplyr::select_if( ~ !all(is.na(.))) %>% 
  dplyr::select(-Comparison)

#Remove transcript Ids
df_one <- df_one %>% 
  tidyr::separate(gene_id, c("gene_id", "transcript")) %>% 
  dplyr::select(-one_of("transcript"))
df_one <- inner_join(df_one, out_group, by = c("gene_id")) %>% 
  group_by(gene_id) %>% 
  dplyr::filter(row_number()==1) %>% 
  tibble::column_to_rownames(var = "gene_id")

conditions <- gtools::combinations(n = ncol(df_one), 2, seq_len(ncol(df_one)))
plot_t(df_one, conditions)
```

```{r heatmapa, fig.width = 8, fig.height = 8 }
colnames(df_one) <- gsub(" t-statistic", "", colnames(df_one))

cor <- cor(df_one)
# clustering based on distance
hcl = hclust(as.dist(1-cor))
# reorder based on clustering
reorder = cor[hcl$order, hcl$order]
results <- pvclust(cor)
# plot clustering
plot(results, hang=-1)
col2 <- colorRampPalette(rev(c("#67001F", "#B2182B", "#D6604D", "#F4A582",
                           "#FDDBC7", "#FFFFFF", "#D1E5F0", "#92C5DE",
                           "#4393C3", "#2166AC", "#053061")))
corrplot(reorder, col = col2(200))
```

# Control vs SCZ
```{r plotb, fig.width = 10, fig.height = 150}
df_two <- df[df$Comparison %in% "Control_vs_SCZ",] %>% 
  dplyr::select_if( ~ !all(is.na(.))) %>% 
  dplyr::select(-Comparison)

#Remove transcript Ids
df_two <- df_two %>% 
  tidyr::separate(gene_id, c("gene_id", "transcript")) %>% 
  dplyr::select(-one_of("transcript"))
df_two <- inner_join(df_two, CMC1, by = c("gene_id")) 
df_two <- inner_join(df_two, CMC_HBCC, by = c("gene_id"))
df_two <- inner_join(df_two, out_group, by = c("gene_id")) %>% 
  group_by(gene_id) %>% 
  dplyr::filter(row_number()==1) %>% 
  tibble::column_to_rownames(var = "gene_id")

conditions <- gtools::combinations(n = ncol(df_two), 2, seq_len(ncol(df_two)))
plot_t(df_two, conditions)
```

```{r heatmapb, fig.width = 8, fig.height = 8 }
colnames(df_two) <- gsub(" t-statistic", "", colnames(df_two))

cor <- cor(df_two)
# clustering based on distance
hcl = hclust(as.dist(1-cor))
# reorder based on clustering
reorder = cor[hcl$order, hcl$order]
results <- pvclust(cor)
# plot clustering
plot(results, hang=-1)
plot(hcl, hang=-1)
corrplot(reorder, col = col2(200))
```

```{r custom.functions}

library(metafor)

# perform meta-analysis given fits from limma
run_meta_analysis = function( tabList, coef, method="REML" ){

  if( length(coef) > 1){
    stop("Only supports one coefficient term")
  }

@@

plotVolcano = function(df, nshow=10){

  library(org.Hs.eg.db)
  library(ggrepel)

  # indicate significant genes
  df$isSignif = c("no","yes")[(df$adj.P.Val < 0.05)+1]

@@

plot_forrest = function( resList, ensID){

  library(patchwork)

  # create a single data.frame with logFC and se
  df_forrest = lapply(resList, function(x)x[ensID,])
  df_forrest = do.call(rbind, df_forrest)
  df_forrest$Dataset = rownames(df_forrest)
  df_forrest$se = df_forrest$logFC /  df_forrest$t

@@

# Violin plots
```{r make.volcano}

df_mpp_dlpfc <- readr::read_tsv(synGet("syn22079472")$path) %>% 
  dplyr::select(Comparison, gene_id, AveExpr, t, logFC, P.Value, adj.P.Val) %>% 
  dplyr::filter(Comparison == 'Control_vs_SCZ') %>% 
  dplyr::rename(`MSSM-Penn-Pitt DLPFC t-statistic` = t)  %>% 
  as.data.frame  %>% unique
rownames(df_mpp_dlpfc) = df_mpp_dlpfc$gene_id

plotVolcano( df_mpp_dlpfc, 15) + ggtitle("MSSM-Penn-Pitt - DLPFC")
```

```{r meta_analysis.SZ}

dlpfcList = list(MPP  = mpp_dlpfc %>% unique %>%
                        dplyr::filter( Comparison == 'Control_vs_SCZ') %>%
                        dplyr::rename(t = `MSSM-Penn-Pitt DLPFC t-statistic`) %>%
                        as.data.frame,
                 HBCC = hbcc_dlpfc %>%  unique %>%
                        dplyr::filter( Comparison == 'Control_vs_SCZ') %>% 
                        dplyr::rename(t = `HBCC DLPFC t-statistic`) %>%
                        as.data.frame)

@@

accList = list(MPP  = mpp_acc %>% unique %>%
                        dplyr::filter( Comparison == 'Control_vs_SCZ') %>%
                        dplyr::rename(t = `MSSM-Penn-Pitt ACC t-statistic` )%>%
                        as.data.frame,
              HBCC  = hbcc_dsACC %>% unique %>%
                        dplyr::filter( Comparison == 'Control_vs_SCZ') %>% 
                        dplyr::rename(t = `HBCC dsACC t-statistic`) %>%
                        as.data.frame)

@@

```{r plot.meta_analysis}

library(cowplot)

table(res_dlpfc$adj.P.Val < 0.05)
table(res_acc$adj.P.Val < 0.05)

# forrest plot of meta-analysis
# This is a bit of a mess
ensID = 'ENSG00000198771.11'
res_dlpfc[ensID,]

lapply(dlpfcList, function(x) x[ensID,])

@@

ensID = rownames(res_dlpfc)[which.min(res_dlpfc$adj.P.Val)][1]

plot_forrest(dlpfcList, ensID)

@@

# Other vs SCZ
```{r plotc, fig.width = 10, fig.height = 100}
df_three <- df[df$Comparison %in% "Other_vs_SCZ",] %>% 
  dplyr::select_if( ~ !all(is.na(.))) %>% 
  dplyr::select(-Comparison)

#Remove transcript Ids
df_three <- df_three %>% 
  tidyr::separate(gene_id, c("gene_id", "transcript")) %>% 
  dplyr::select(-one_of("transcript"))
df_three <- inner_join(df_three, out_group, by = c("gene_id")) %>% 
  group_by(gene_id) %>% 
  dplyr::filter(row_number()==1) %>% 
  tibble::column_to_rownames(var = "gene_id")

conditions <- gtools::combinations(n = ncol(df_three), 2, seq_len(ncol(df_three)))
plot_t(df_three, conditions)
```

```{r heatmapc, fig.width = 8, fig.height = 8 }
colnames(df_three) <- gsub(" t-statistic", "", colnames(df_three))

cor <- cor(df_three)
# clustering based on distance
hcl = hclust(as.dist(1-cor))
# reorder based on clustering
reorder = cor[hcl$order, hcl$order]
results <- pvclust(cor)
# plot clustering
plot(results, hang=-1)
corrplot(reorder, col = col2(200))
```

Code: r thisFile

```{r store.files, eval=FALSE}
CODE <- Folder(name = "Sex-specific Meta-analysis", parentId = "syn21435176")
CODE <- synStore(CODE)
```

>>>>

---
title: "Meta-analysis on diagnosis in MSSM-Penn-Pitt and HBCC ACC/DLPFC"
author: "KM and GH"
date: "`r date()`"
output: html_document
---

```{r knit2synapse, eval=FALSE}
library(synapser)
library(knit2synapse)
synLogin()
knit2synapse::createAndKnitToFolderEntity(file = "individual-cohorts/meta-analysis.Rmd",
                                          parentId ="syn22276904",
                                          folderName = 'Meta-analysis on diagnosis')
```

```{r setup, include=FALSE}
library(synapser)

@@

library(ggplot2)
library(dplyr)
library(ComplexHeatmap)
library(readxl)
library(gtools)
library(corrplot)
library(pvclust)
library(metafor)
library(cowplot)
library(org.Hs.eg.db)
library(ggrepel)
library(metafor)
library(patchwork)

knitr::opts_chunk$set(
  echo = FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=FALSE, eval=TRUE}
thisFileName <- 'meta-analysis.Rmd'

@@

```{r input_data}
mpp_acc <- readr::read_tsv(synGet("syn22079117")$path) %>% 
  dplyr::select(Comparison, gene_id, AveExpr, t, logFC, P.Value, adj.P.Val)
mpp_dlpfc <- readr::read_tsv(synGet("syn22079472")$path) %>% 
  dplyr::select(Comparison, gene_id, AveExpr, t, logFC, P.Value, adj.P.Val)
hbcc_dsACC <- readr::read_tsv(synGet("syn22086829")$path) %>% 
  dplyr::select(Comparison, gene_id, AveExpr, t, logFC, P.Value, adj.P.Val) 
hbcc_dlpfc <- readr::read_tsv(synGet("syn22084205")$path) %>% 
  dplyr::select(Comparison, gene_id, AveExpr, t, logFC, P.Value, adj.P.Val)
```

@@

```{r custom.functions}
# perform meta-analysis given fits from limma
run_meta_analysis = function(tabList, method="FE") {

@@

plotVolcano = function(df, nshow=10){
  # indicate significant genes
  df$isSignif = c("no","yes")[(df$adj.P.Val < 0.05)+1]

@@

plot_forrest = function(resList, ensID) { 
  # create a single data.frame with logFC and se
  df_forrest = lapply(resList, function(x) if(ensID %in% x[ensID,]) {
    x[ensID,]
  } 
  )
  df_forrest[sapply(df_forrest, is.null)] <- NULL
  df_forrest = do.call(rbind, df_forrest)
  df_forrest$Dataset = rownames(df_forrest)
  df_forrest$se = df_forrest$logFC /  df_forrest$t

@@

# Control vs SCZ
```{r meta_analysis.SZ}
make_list <- function(mpp, hbcc, string) {
  list <- list(MPP = mpp %>%
                 unique() %>% 
                 dplyr::filter(Comparison == string) %>% 
                 as.data.frame,
               HBCC = hbcc %>% 
                 unique() %>% 
                 dplyr::filter(Comparison == string) %>% 
                 as.data.frame)
}

dlpfcList <- make_list(mpp_dlpfc, hbcc_dlpfc, string = "Control_vs_SCZ")
accList <- make_list(mpp_acc, hbcc_dsACC, string = "Control_vs_SCZ")

@@

@@

```{r plot.meta_analysisSCZ}
kable(table(res_dlpfc$adj.P.Val < 0.05), col.names = c("", "Num. genes with p-value < 0.05"))
kable(table(res_acc$adj.P.Val < 0.05), col.names = c("", "Num. genes with p-value < 0.05"))

@@

@@

# Control vs Other/SCZ
```{r meta_analysis.SCZOther}
make_list <- function(mpp, hbcc, string) {
  list <- list(MPP = mpp %>%
                 unique() %>% 
                 dplyr::filter(Comparison == string) %>% 
                 as.data.frame,
               HBCC = hbcc %>% 
                 unique() %>% 
                 dplyr::filter(Comparison == string) %>% 
                 as.data.frame)
}

dlpfcList <- make_list(mpp_dlpfc, hbcc_dlpfc, string = "Control_vs_OtherSCZ")
accList <- make_list(mpp_acc, hbcc_dsACC, string = "Control_vs_OtherSCZ")

```{r plot.meta_analysisSCZOther}
kable(table(res_dlpfc$adj.P.Val < 0.05), col.names = c("", "Num. genes with p-value < 0.05"))
kable(table(res_acc$adj.P.Val < 0.05), col.names = c("", "Num. genes with p-value < 0.05"))

# Control vs Other
```{r meta_analysis.CTOther}
make_list <- function(mpp, hbcc, string) {
  list <- list(MPP = mpp %>%
                 unique() %>% 
                 dplyr::filter(Comparison == string) %>% 
                 as.data.frame,
               HBCC = hbcc %>% 
                 unique() %>% 
                 dplyr::filter(Comparison == string) %>% 
                 as.data.frame)
}

dlpfcList <- make_list(mpp_dlpfc, hbcc_dlpfc, string = "Control_vs_Other")
accList <- make_list(mpp_acc, hbcc_dsACC, string = "Control_vs_Other")

```{r plot.meta_analysisCTOther}
kable(table(res_dlpfc$adj.P.Val < 0.05), col.names = c("", "Num. genes with p-value < 0.05"))
kable(table(res_acc$adj.P.Val < 0.05), col.names = c("", "Num. genes with p-value < 0.05"))

# Other vs SCZ
```{r meta_analysis.OtherSCZ}
make_list <- function(mpp, hbcc, string) {
  list <- list(MPP = mpp %>%
                 unique() %>% 
                 dplyr::filter(Comparison == string) %>% 
                 as.data.frame,
               HBCC = hbcc %>% 
                 unique() %>% 
                 dplyr::filter(Comparison == string) %>% 
                 as.data.frame)
}

dlpfcList <- make_list(mpp_dlpfc, hbcc_dlpfc, string = "Other_vs_SCZ")
accList <- make_list(mpp_acc, hbcc_dsACC, string = "Other_vs_SCZ")

```{r plot.meta_analysisOtherSCZ}
kable(table(res_dlpfc$adj.P.Val < 0.05), col.names = c("", "Num. genes with p-value < 0.05"))
kable(table(res_acc$adj.P.Val < 0.05), col.names = c("", "Num. genes with p-value < 0.05"))

## RCSD1

Forest plot of example top gene.

```{r specific-gene}
# forrest plot of meta-analysis
ensID = 'ENSG00000198771.11'
kable(res_dlpfc[ensID,])
lapply(dlpfcList, function(x) x[ensID,])
ensID = rownames(res_dlpfc)[which.min(res_dlpfc$adj.P.Val)][1]
plot_forrest(dlpfcList, ensID)
```

### Source
